Trans-Dimensional Inference for Gravitational Lens Substructures

### 2015-05-18 Papers/ms.tex
citep, not ref
Lists of parameters

The differences between the approaches can be broken down into
i) different assumptions about the prior information being used, and ii)
different computational methods used to compute the results.
Each approach involves tradeoffs between convenience and realism.
Simply parameterized models, such as Sérsic surface
brightness profiles for the source, and SIE models for the lens, are very
convenient. They only have a few adjustable parameters, and they capture
(to ``first order'') relevant prior information that we have about the
source and lens profiles. Of course, they are clearly simplifications,
and can produce misleading results if the actual profile is very different
from any member of the assumed family.

On the
other hand, pixellated models for the source \citep[e.g.][]{suyu} or the lens
\citep[e.g.][]{2014MNRAS.445.2181C} can in principle represent ``any''
source surface brightness profile or lens density profile. However, the prior
distribution over pixel values is often chosen to be a multivariate gaussian
for mathematical reasons, so
that the source can be analytically marginalized over
\citep{2003ApJ...590..673W}. Unfortunately, a multivariate gaussian prior
over pixel intensities can correspond to a very poor model of our prior
beliefs about the source. It assigns virtually zero prior probability
to the hypothesis that the source actually looks like a galaxy, and very high
prior probability to the hypothesis that the source looks like noise (or
blurred noise).

In a previous paper \citep{2011MNRAS.412.2521B} we argued that an ideal
modelling approach lies somewhere between simply-parameterised and
pixellated models. The source and the lens can be built up from a moderate
number of simply parameterised components. For the source, this allows
us to incorporate prior knowledge about the local correlations (the surface
brightness at any particular point is likely to be similar to that at a
nearby point) and the fact that most of the sky is dark.

For the lens, if we are interested in detecting and measuring the properties of
dark substructures, a superposition of an unknown number of ``blobs'' is the
relevant model. In addition, if we want to constrain the ``mass function'' of
these blobs, we will need a hierarchical model which specifies the probability
distribution for the masses, given some hyperparameters.

@@

\subsection{The Source}
The surface brightness profile of the source is assumed to be composed of
a sum of a finite number of ``blobs'', or circular basis functions.
The surface brightness of a single blob centered at the origin
is:
\begin{eqnarray}
f(x, y) &=& \left\{
\begin{array}{lr}
\frac{A}{2\pi w^2}\left[1 - \left(\frac{r}{w}\right)^2\right], & r \leq w\\
0, & \textnormal{otherwise}
\end{array}
\right.
\end{eqnarray}
where $r = \sqrt{x^2 + y^2}$, $w$ is the width of the blob, and $A$ is the
total flux of the blob (i.e. the integral of the surface brightness).
These basis functions are faster to evaluate than gaussians, since they do not
contain an exponential function. In addition, the finite support means that
each blob will evaluate to zero over a large fraction of the domain which
confers an additional speed advantage.

@@

@@

Since the model is not of fixed dimension (the number of source and lens
``blobs'' is unknown), we implemented the MCMC using the {\tt RJObject}
\citep{rjobject} software, a C++ library for implementing trans-dimensional
MCMC with hierarchically-specified priors, and doing the sampling with the
Diffusive Nested Sampling algorithm \citep{dnest}. Previous astronomical
applications of {\tt RJObject} include \citet{magnetron}
and~\ref{exoplanet}.

>>>>

The differences between the approaches can be broken down into
i) different assumptions about the prior information being used, and ii)
different computational methods used to compute the results.
Each approach involves tradeoffs between convenience and realism.
Simply parameterized models, such as Sérsic surface
brightness profiles for the source, and singular isothermal ellipsoid plus
external shear (SIE+$\gamma$) models for the lens, are very
convenient. They only have a few adjustable parameters, and they capture
(to ``first order'') relevant prior information that we have about the
source and lens profiles. Of course, they are clearly simplifications,
and can produce misleading results if the actual profile is very different
from any member of the assumed family.

On the
other hand, pixellated models for the source \citep[e.g.][]{suyu} or the lens
\citep[e.g.][]{2014MNRAS.445.2181C} can in principle represent ``any''
source surface brightness profile or lens projected density profile. However, the prior
distribution over pixel values is often chosen to be a multivariate gaussian
for mathematical reasons, so
that the source can be analytically marginalized over
\citep{2003ApJ...590..673W}. Unfortunately, a multivariate gaussian prior
over pixel intensities can correspond to a very poor model of our prior
beliefs about the source. It assigns virtually zero prior probability
to the hypothesis that the source actually looks like a galaxy, and very high
prior probability to the hypothesis that the source looks like noise (or
blurred noise). These priors also assign nonzero probability to negative
surface brightness or density values.

\citet{2011MNRAS.412.2521B} argued that an ideal
modelling approach lies somewhere between simply-parameterised and
pixellated models. The source and the lens can be built up from a moderate
number of simply parameterised components. For the source, this allows
us to incorporate prior knowledge about the local correlations (the surface
brightness at any particular point is likely to be similar to that at a
nearby point) and the fact that most of the sky is dark.

For the lens, if we are interested in detecting and measuring the properties of
dark substructures, a superposition of an unknown number of ``blobs'' is the
most natural model. In addition, if we want to constrain the ``mass function'' of
these blobs, we will need a hierarchical model which specifies the probability
distribution for the masses, conditional on some hyperparameters. Inferring the
mass function of the blobs then reduces to calculating the posterior
distribution for the hyperparameters.

@@

\subsection{The Source}
The surface brightness profile of the source is assumed to be composed of
a sum of a finite number of ``blobs'', or basis functions. The specific choice
of basis functions we use is motivated primarily by computational speed.
The surface brightness of a single blob centered at the origin
is:
\begin{eqnarray}
f(x, y) &=& \left\{
\begin{array}{lr}
\frac{A}{2\pi w^2}\left[1 - \left(\frac{r}{w}\right)^2\right], & r \leq w\\
0, & \textnormal{otherwise}
\end{array}
\right.
\end{eqnarray}
where $r = \sqrt{x^2 + y^2}$, $w$ is the width of the blob, and $A$ is the
total flux of the blob (i.e. the integral of the surface brightness over the
entire domain).
These basis functions are faster to evaluate than gaussians, since they do not
contain an exponential function. In addition, the finite support means that
each blob will evaluate to zero over a large fraction of the domain which
confers an additional speed advantage.

@@

Under these assumptions, the source can be described by the following
parameters:

\begin{eqnarray}
\left\{
N_{\rm src}, \left\{\mathbf{x}^{\rm src}_i\right\}_{i=1}^N
\right\}
\end{eqnarray}
where $\mathbf{x}^{\rm src}_i$ denotes the parameters for source blob $i$:
\begin{eqnarray}
\mathbf{x}^{\rm src}_i &=& \left\{x_i, y_i, A_i, w_i
\right\}.
\end{eqnarray}

@@

Since the model is not of fixed dimension (the number of source and lens
``blobs'' is unknown), we implemented the MCMC using the {\tt RJObject}
\citep{rjobject} software, a C++ library for implementing trans-dimensional
MCMC with hierarchically-specified priors, and doing the sampling with the
Diffusive Nested Sampling algorithm \citep{dnest}. Previous astronomical
applications of {\tt RJObject} include \citet{magnetron}
and~\citep{exoplanet}.
